Gusto: a toolkit for compatible finite element dynamical cores

### 2025-08-06 source/paper.tex
add pseudocode for timesteppers

\subsection{Viscous and diffusive terms}

For the simulations presented here we do not require viscous or
diffusive terms for stability, but sometimes they are required to
produce converged solutions so that refining the grid does not
lead to smaller and smaller scale motions. Here we present the symmetric
interior penalty discretisation method used for these terms in Gusto.

@@

\JScomment{This paragraph should refer to the schematic and explain
  all aspects} The Gusto model comprises a domain and a prognostic
equation set. The domain holds the mesh and the finite element
function spaces along with any other information about the space-time
domain such as the coordinates and the timestep. The prognostic
equation sets are each defined in a python class, making use of the
common finite element forms for the transport and diffusion terms to
set up the residual (defined in section \ref{}) in weak form. At this
point, no particular discretisation of the transport or diffusion
forms is specified. This happens in the \texttt{SpatialMethod} class
which can be used to apply the methods described in section \ref{} to
the relevant terms. Any physics source terms, such as **, are added by
calling the \texttt{Physics} class that defines them. The
\texttt{TimeStepper} class contains the definition of the timestepping
procedure and the \texttt{run} method that loops over the
timesteps. It also sets up diagnostic fields and calls the IO as
required. The simplest timeloop class will just apply one time
discretisation method to the equation set but it is common in weather
and climate prediction models to split the different types of terms
and apply a time discretisation separately to each type of term. For
example, the transport and forcing terms might be solved for
separately, such as in **, which enables explicit transport schemes to
be applied to each field. ** something about SIQN structure. Terms
involved in physics parameterisations are evaluated in the appropriate
part of the iteration loop, with physics processes that occur on
`fast' timescales evaluated more frequently and `slower' physics
processes evaluated just once at the end of the timestep. The choice
of where to evaluate the physics processes in the iteration is also
based on the computational cost and the best way to do this is still
an open research question (**ref).

\begin{itemize}
\item what about initialisation and diagnostics?
\end{itemize}

\begin{algorithm}
  \caption{Pseudocode for SIQN timestep loop}\label{alg:SIQN}
  \begin{algorithmic}
    \STATE
  \end{algorithmic}
\end{algorithm}

\begin{itemize}
\item generic example(s) - no FML, just pseudocode for basic
  timestepper, SIQN and split physics
\item lead in to next section... facilitated by FML i.e. timesteppers don't care what equations they're given because forms are labelled
\end{itemize}

>>>>

\subsection{Viscous and diffusive terms}
\label{sec: diffusion}
For the simulations presented here we do not require viscous or
diffusive terms for stability, but sometimes they are required to
produce converged solutions so that refining the grid does not
lead to smaller and smaller scale motions. Here we present the symmetric
interior penalty discretisation method used for these terms in Gusto.

@@

\JScomment{This paragraph should refer to the schematic and explain
  all aspects} The Gusto model comprises a domain and a prognostic
equation set. The domain holds the mesh and the finite element
function spaces along with any other information about the space-time
domain such as the coordinates and the timestep. The prognostic
equation sets are each defined in a python class, making use of the
common finite element forms for the transport and diffusion terms to
set up the residual (i.e. the left hand side of the equations defined
in section \ref{sec: equations}) in weak form. At this point, no
particular discretisation of the transport or diffusion forms is
specified. This happens in the \texttt{SpatialMethod} class which can
be used to apply the methods described in sections \ref{sec:
  transport} and \ref{sec: diffusion} to the relevant terms. Any
physics source terms, such as those required for moist processes or
tubulence parameterisation, are added by calling the \texttt{Physics}
class that defines them. The \texttt{TimeStepper} class contains the
definition of the timestepping procedure and the \texttt{run} method
that loops over the timesteps. It also sets up diagnostic fields and
calls the IO as required. The simplest timeloop class will just apply
one time discretisation method to the equation set but it is common in
weather and climate prediction models to split the different types of
terms and apply a time discretisation separately to each type of
term. For example, the transport and forcing terms might be solved for
separately, such as in **, which enables explicit transport schemes to
be applied to each field. ** something about SIQN structure. Terms
involved in physics parameterisations are evaluated in the appropriate
part of the iteration loop, with physics processes that occur on
`fast' timescales evaluated more frequently and `slower' physics
processes evaluated just once at the end of the timestep. Since the
test cases presented later only use simple moist physics
parameterisations, we just include the option to solve for these at
the end of the timestep.

\begin{algorithm} 
  \caption{Basic timestepping scheme}
  \label{alg: basic}
  \begin{algorithmic}[1]
    \STATE Replace transport and diffusion terms with specified spatial discretisation scheme.
    \STATE Setup IO and diagnostics and write out initial conditions.
    \WHILE{$t < t_{max}$}
      \STATE Apply timestepping scheme to equation set: $\chi^{n+1} = \mathcal{T}(\chi^n, \chi^{n+1})$
      \STATE Advance timestep: $\chi^n = \chi^{n+1}$
      \IF{ndt \% nts == 0}
        \STATE Update timeseries diagnostics file.
      \ENDIF
      \IF{ndt \% ndiag == 0}
        \STATE Compute and write out diagnostic fields.
      \ENDIF
      \IF{ndt \% nchk == 0}
        \STATE Write out checkpoint file.
      \ENDIF
    \ENDWHILE
  \end{algorithmic} 
\end{algorithm}

\begin{algorithm} 
  \caption{Split physics timestepping scheme}
  \label{alg: split physics}
  \begin{algorithmic}[1]
    \STATE Dynamics step: $\chi^{n+1} = \mathcal{D}(\chi^n)$
    \FOR {$\mathcal{P}$ in physics schemes}
      \STATE Apply physics scheme: $\chi^{n+1} = \chi^{n+1} + \mathcal{P}(\chi^{n+1})$
    \ENDFOR
    \STATE Advance timestep: $\chi^n = \chi^{n+1}$
  \end{algorithmic} 
\end{algorithm}

\begin{algorithm} 
  \caption{Semi-implicit quasi-Newton timestepping scheme}
  \label{alg: SIQN}
  \begin{algorithmic}[1]
    \STATE Set: $\chi^{n+1} = \chi^{n}$
    \STATE Explicit forcing: $\chi_{f, e} = \chi^n + \frac{1}{2}\Delta t \mathcal{F}(\chi^n)$
    \FOR {$OUTER$}
      \STATE Update advecting velocity: $\bar{\MM{u}} = \frac{1}{2} (\MM{u}^{n+1} + \MM{u}^n)$
      \STATE Explicit transport: $\chi_T = \mathcal{A}_{\bar{\MM{u}}}(\chi_{f,e})$
      \FOR {$INNER$}
        \STATE Implicit forcing: $\chi_{f,i} = \chi_T + \frac{1}{2}\Delta t \mathcal{F}(\chi^{n+1})$
        \STATE Write residual: $\chi_{r} = \chi_{f,i} - \chi^{n+1}$
        \STATE Solve: $\mathcal{S}(\chi') = \chi_{r}$ for $\chi'$
        \STATE Increment: $\chi^{n+1} = \chi^{n+1} + \chi'$
      \ENDFOR
    \ENDFOR
    \STATE Explicit physics: $\chi^{n+1} = \mathcal{P}(\chi^{n+1})$
    \STATE Advance timestep: $\chi^n = \chi^{n+1}$
  \end{algorithmic} 
\end{algorithm}
